Hijacking Simulators with Universal Probabilistic Programming

### 2019-08-15 paper/learning/notes.tex
save plots
adding SVI notes and example - still to finish

Here the rejection sampling functions take an additional argument $1$, or $0$,
to specify whether or not we want to collect samples to train, $1$, else replace with 
the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
the unknown densities that we wish to learn through an amortized approach.
It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
\begin{equation}
\text{derivation}
\end{equation}

@@

\section{An Example: Modelling Parasite Densities in Simulation}
% \begin{wrapfigure}{r}{0.4\textwidth}
  \begin{figure}[h!]
    \centering
    \label{fig:parasite}
    \includegraphics[width=0.4\textwidth]{parasite_model.pdf}
    \caption{The DAG of the Parasite model.}
  % \end{wrapfigure}
  \end{figure}
The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
model for determining the number of infections within a given population demographic.
The models hypothesized in ~\cite{smith2006relationship} were found, 
after extensive testing and development, to not characterise well the number 
of infections among different demographics, especially those under 15 years old. 
By using probabilistic programming we can immediately understand 
the expressiveness of the proposed model and can exact important insights for 
extending the proposed models, which increases ones ability to develop 
improved models to estimate parasite densities, given 
our observations of the seasonal entomological inoculation rate (EIR).
In addition to this we can greatly reduce the computational costs, whilst
improving model comprehension and expressivity.

>>>>

Here the rejection sampling functions take an additional argument $1$, or $0$,
to specify whether or not we want to collect samples to train, $1$, else replace with 
the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
the unknown densities that we wish to learn through an amortized approach.
It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
\textbf{Bradley:  You only arrive at this, if the density is uniform}
\begin{align*}
P( X= x) = \sum^{\infty}_{n=1} P(\text{reject $n-1$ times}, \text{draw} Y = x) \\
\text{todo} 
\end{align*}

@@

\section{Stochastic Variational Inference}

Using the standard notations we define the joint as follows:
\begin{equation}
  p(x, z ; \theta) = p_{\theta}(x,z) = p_{\theta}(x| z)p_{\theta}(z)
\end{equation}

We assume that the distributions that create the \emph{model} $p_{\theta}(x, z)$ can be:
\begin{enumerate}
  \item We can sample from each $p_i$
  \item We can compute the pointwise log pdf of $p_i$
  \item $p_i$ is differentiable w.r.t the parameters $\theta$
  \end{enumerate}

\subsection{Learning the model}

Our objective will either want to maximize, or minimize, depending on what we want to discover.

\begin{equation}
  \theta_{\text{min} \backslash \text{max} } = \text{argmax$\backslash$min}_{\theta}\log p_\theta(x)
\end{equation}

where the log evidence $p_\theta(x)$ is given by:
\begin{equation}
  \log p_{\theta}(x) = \log \int dz p_\theta(x,z)
\end{equation}

This is a doubly hard problem as even if $\theta$ is fixed the integral over the set
of latent variables defined by $z$ is intractable. 
Even if we know how to calculate the log evidence for all $\theta$, maximising or minimising, 
the log evidence as a function of $\theta$ is in general a difficult non-convex problem.

In addition to finding the parameters $\theta$ we also require the posterior over
the latent variables $z$:
\begin{equation}
  p_{\theta_{\text{max$\backslash$min}}}(z|x) \frac{p_{\theta}(x,z)}{\int dz p_{\theta_{\text{max$\backslash$min}}}(x,z)} 
\end{equation}

To this in the variational inference setting we learn an approximation to $p_{\theta_{\text{max$\backslash$min}}}(z|x)$
denoted by $q_\phi(z)$ where $\phi$ are the variational parameters.

\section{The Objective}

In many of these settings the ELBO loss characterises the problem that we 
aim to solve.

\begin{equation}
  ELBO = \mathbb{E}_{q_{\phi}(z)} [ \log p_{\theta}(x,z) - \log q_{\phi}(z) ]
\end{equation}
and we have for all choices of $\theta$ and $\phi$ that:
\begin{equation}
  \log p_{\theta}(x) \geq ELBO 
\end{equation}
\section{An Example: Modelling Parasite Densities in Simulation}
% \begin{wrapfigure}{r}{0.4\textwidth}
  \begin{figure}[h!]
    \centering
    \label{fig:parasite}
    \includegraphics[width=0.4\textwidth]{parasite_model.pdf}
    \caption{The DAG of the Parasite model.}
  % \end{wrapfigure}
  \end{figure}
The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
model for determining the number of infections within a given population demographic.
The models hypothesized in ~\cite{smith2006relationship} were found, 
after extensive testing and development, to not characterise well the number 
of infections among different demographics, especially those under 15 years old. 
By using probabilistic programming we can immediately understand 
the expressiveness of the proposed model and can exact important insights for 
extending the proposed models, which increases ones ability to develop 
improved models to estimate parasite densities, given 
our observations of the seasonal entomological inoculation rate (EIR).
In addition to this we can greatly reduce the computational costs, whilst
improving model comprehension and expressivity.
